---
title: "hellometry"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{hellometry}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---
```{r include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  warning = FALSE,
  message = FALSE)
```
## Introduction to package use
Welcome to hellometry! This simple, lightweight package allows you to quickly
impute missing body size and body mass measurements in the supplied data frame (or tibble). The package uses existing measurements in your data to estimate the missing ones, so, essentially, the better the data supplied the better the estimations will be. It is compatible with any kind of data, provided you follow a few simple rules of column naming. These are as follow: <br>
- column with measurements needs to be called "size_col", and only accepts numerical measurements or "unknown", "small", "medium" or "large" as categorical values <br>
- column with development stage of specimens (larva/pupa/adult for instance) needs to be called "stage" <br>
- column with abundance of specimens needs to be called "abundance" <br>
- column with biomass of specimens needs to be called 'biomass_col'

The idea of the package, wrapped in `hellometry()` is rather simple. First, a `measurement_table` is compiled from your own measurements. This table is then used to compute every possible size estimations, and allometric linear models with  `full_estimation_table()`. Models are  deleted if p > 0.05 or R2 > 0.95, unless you specify other values. Note that size and models are not necessarily computed at the same scale, it all depends on model fit filtered by these p and R2 values! Finally, the size estimation and model at the lowest possible taxonomic level are joined back to the data, and used to estimate biomass at that particular row. The estimations are given with a prediction interval, which differs from a confidence interval in that it represents uncertainty around a single value as opposed to mean predicted values. `hellometry()` returns a list of three elements: your data with estimated body sizes and biomasses, and column with information on the level at which were performed the estimations, a tibble with all unique size estimations that were joined to your data, and a tibble with all unique allometric models that were joined to your data. <br>

The diagram below traces one example species (`Sp.A`) through every stage of the pipeline, from your raw input to the final estimated table.

```
                       hellometry() PIPELINE
                       =====================

  [1] YOUR DATA                          size_col may be numeric OR a
  ┌─────────┬───────┬──────────┬─────────────┐   category: small / medium /
  │ species │ stage │ size_col │ biomass_col │   large / unknown.
  ├─────────┼───────┼──────────┼─────────────┤   NA biomass_col = to estimate.
  │ Sp.A    │ larva │   3.8    │    12.1     │
  │ Sp.A    │ larva │   5.9    │    15.2     │
  │ Sp.A    │ larva │   12     │    24.3     │
  │ Sp.A    │ larva │ unknown  │    NA       │  <- to estimate
  │ Sp.A    │ larva │   8.1    │    17.2     │
  │ Sp.A    │ larva │   8.2    │    NA       │
  │  ...    │  ...  │   ...    │    ...      │
  └─────────┴───────┴──────────┴─────────────┘
        │
        │  make_measurement_table()  ->  keep rows with a numeric size,
        v                                 one row per individual (abundance)
  [2] MEASUREMENT TABLE
  ┌─────────┬───────┬──────────┬─────────────┐
  │ species │ stage │ size_col │ biomass_col │
  ├─────────┼───────┼──────────┼─────────────┤
  │ Sp.A    │ larva │   3.8    │    12.1     │
  │ Sp.A    │ larva │   5.9    │    15.2     │
  │ Sp.A    │ larva │   12     │    24.3     │
  │ Sp.A    │ larva │   8.1    │    17.2     │
  │ Sp.A    │ larva │   8.2    │    NA       │
  │  ...    │  ...  │   ...    │    ...      │
  └─────────┴───────┴──────────┴─────────────┘
        │
        │   full_estimation_table()  --  computed at every taxonomic level
        ├───────────────────────────────┬───────────────────────────────┐
        v what = "size_col"             v what = "biomass_col"
  [3a] SIZE ESTIMATIONS                 [3b] BIOMASS MODELS
  ┌─────────┬──────┬───────┬───────────────┬──────┐  ┌─────────┬──────┬───────┬───────┐
  │ level   │ name │ stage │ size_category │ size │  │ level   │ name │ stage │ model │
  ├─────────┼──────┼───────┼───────────────┼──────┤  ├─────────┼──────┼───────┼───────┤
  │ species │ Sp.A │ larva │ unknown       │ 7.6  │  │ species │ Sp.A │ larva │  lm   │
  │ species │ Sp.A │ larva │ small         │ 4.85 │  │ genus   │ Gn.1 │ larva │  lm   │
  │ species │ Sp.A │ larva │ medium        │ 8.15 │  │ family  │ Fm.1 │ larva │  lm   │
  │ species │ Sp.A │ larva │ large         │ 12   │  │  ...    │ ...  │  ...  │  ...  │
  │ genus   │ Gn.1 │ larva │ unknown       │ 6.9  │  └─────────┴──────┴───────┴───────┘
  │  ...    │ ...  │  ...  │   ...         │ ...  │        │
  └─────────┴──────┴───────┴───────────────┴──────┘        │
        │                                                  │
        │  join finest-level size back to data             │
        v                                                  │
  [4] SIZES FILLED                                         │
  ┌─────────┬───────┬──────────┬─────────────┬────────────┐│
  │ species │ stage │ size_col │ biomass_col │ size_level ││
  ├─────────┼───────┼──────────┼─────────────┼────────────┤│
  │ Sp.A    │ larva │   3.8    │    12.1     │ raw_data   ││
  │ Sp.A    │ larva │   7.6    │    NA       │ species    ││  <- "unknown" -> 7.6
  │  ...    │  ...  │   ...    │    ...      │   ...      ││
  └─────────┴───────┴──────────┴─────────────┴────────────┘│
        │                                                  │
        │  apply finest-level model: biomass = predict(lm, size) <--┘
        v
  [5] FINAL OUTPUT  ($data)
  ┌─────────┬───────┬──────────┬─────────────┬────────────┬─────────────┐
  │ species │ stage │ size_col │ biomass_col │ size_level │ model_level │
  ├─────────┼───────┼──────────┼─────────────┼────────────┼─────────────┤
  │ Sp.A    │ larva │   3.8    │    12.1     │ raw_data   │ raw_data    │
  │ Sp.A    │ larva │   7.6    │    16.4     │ species    │ genus       │  <- both estimated
  │ Sp.A    │ larva │   8.2    │    17.5     │ raw_data   │ genus       │  <- mass estimated
  │  ...    │  ...  │   ...    │    ...      │   ...      │    ...      │
  └─────────┴───────┴──────────┴─────────────┴────────────┴─────────────┘

  Legend: raw_data = measured value kept as-is | species/genus/family =
          taxonomic level at which the estimate was made (finest wins).
          $data also carries biomass_lower / biomass_upper (prediction
          interval) and *_taxon_name columns, omitted here for clarity.
```



## Quick working example
The package estimates sizes and biomasses from the measurements you provide.
To illustrate, let us generate a random dataset of reptiles. <br>
```{r}
# Let's generate a random dataset of reptiles
set.seed(123) # For reproducibility

# Define orders, families, genera, and species
## Orders
orders <-
  c("Squamata", "Testudines", "Crocodilia", "Rhynchocephalia")
## Families
families <- list(
  Squamata = c("Colubridae", "Viperidae", "Elapidae", "Scincidae"),
  Testudines = c("Testudinidae", "Emydidae", "Chelydridae"),
  Crocodilia = c("Crocodylidae", "Alligatoridae"),
  Rhynchocephalia = c("Sphenodontidae")
)
## Genera
genera <- list(
  Colubridae = c("Coluber", "Lampropeltis", "Pantherophis"),
  Viperidae = c("Crotalus", "Vipera"),
  Elapidae = c("Naja", "Bungarus"),
  Scincidae = c("Scincus", "Eumeces"),
  Testudinidae = c("Gopherus", "Aldabrachelys"),
  Emydidae = c("Trachemys", "Terrapene"),
  Chelydridae = c("Chelydra", "Macrochelys"),
  Crocodylidae = c("Crocodylus", "Osteolaemus"),
  Alligatoridae = c("Alligator", "Caiman"),
  Sphenodontidae = c("Tuatara")
)
## Species
species <- list(
  Coluber = c("Coluber constrictor", "Coluber viridiflavus"),
  Lampropeltis = c("Lampropeltis getula", "Lampropeltis triangulum"),
  Pantherophis = c("Pantherophis guttatus", "Pantherophis obsoletus"),
  Crotalus = c("Crotalus horridus", "Crotalus atrox"),
  Vipera = c("Vipera berus", "Vipera aspis"),
  Naja = c("Naja naja", "Naja pallida"),
  Bungarus = c("Bungarus caeruleus", "Bungarus multicinctus"),
  Scincus = c("Scincus scincus", "Scincus mitranus"),
  Eumeces = c("Eumeces schneideri", "Eumeces egregius"),
  Gopherus = c("Gopherus agassizii", "Gopherus polyphemus"),
  Aldabrachelys = c("Aldabrachelys gigantea"),
  Trachemys = c("Trachemys scripta", "Trachemys gaigeae"),
  Terrapene = c("Terrapene carolina", "Terrapene ornata"),
  Chelydra = c("Chelydra serpentina"),
  Macrochelys = c("Macrochelys temminckii"),
  Crocodylus = c("Crocodylus niloticus", "Crocodylus acutus"),
  Osteolaemus = c("Osteolaemus tetraspis"),
  Alligator = c("Alligator mississippiensis"),
  Caiman = c("Caiman crocodilus"),
  Tuatara = c("Sphenodon punctatus")
)

# Generate empty dataset
myreptiles <-
  data.frame(order = character(),
             family = character(),
             genus = character(),
             species = character(),
             size_col = numeric(),
             biomass_col = numeric(),
             abundance = integer())


# Make loop to fill the dataset at random
## Keeping taxonomic consistency
for (order in orders) {
  for (family in families[[order]]) {
    for (genus in genera[[family]]) {
      for (spec in species[[genus]]) {
        # Generate at least 3 rows per species
        n_rows <-
          sample(3:10, 1)
        # Create a temporary data frame for the current species
        temp <- data.frame(
          order = order,
          family = family,
          genus = genus,
          species = spec,
          # Generate random size and biomass values
          size_col = round(rnorm(n_rows, mean = runif(1, 50, 500), sd = 30), 1),
          biomass_col = round(rnorm(n_rows, mean = runif(1, 100, 5000), sd = 100), 1),
          # Generate random abundance values
           abundance = sample(1:10, n_rows, replace = TRUE))
        ## Add the temporary data to the main dataset
        myreptiles <-
          dplyr::bind_rows(myreptiles,
                           temp)}}}}

# Have a look at the data
dplyr::glimpse(myreptiles)

# Now make a random half "unknown" and NA to estimate
## Determine the number of rows to turn into NA
num_na <-
  floor(nrow(myreptiles) / 2)
# Randomly select rows to have modified values in size_col and biomass_col
na_rows <-
  sample(1:nrow(myreptiles), num_na)
## Assign new values to the selected rows in the specified columns
myreptiles$size_col[na_rows] <-
  "unknown"
myreptiles$biomass_col[na_rows] <-
  NA
## Now we should see altered values
dplyr::glimpse(myreptiles)


# Now we can use the package to estimate sizes and biomasses!

# Load library
library(hellometry)

# We need to define a suite of levels to be used for estimations, a.k.a. columns in the data
level_vec  <-
  c("species", "genus", "family", "order")

# The data also need to include a biomass_type (dry/wet) column, here use dry because it tends to be more precise, and a stage column for organisms with very different life stages, here put adult because it does not really matter
myreptiles <-
  myreptiles %>%
  dplyr::mutate(biomass_type = "dry") %>%
  dplyr::mutate(stage = "adult")

# Get the estimations
myreptiles_estimated <-
  hellometry(dats = myreptiles, ## The data to be used
              level_vec  = level_vec , ## The taxonomic levels to be used
              biomass_type = "dry") ## Type of biomass to be estimated, "dry" or "wet"

# The result is a list of three elements
# - `dats` - your data with estimated body sizes and biomasses, and column with information on the level at which were performed the estimations
dplyr::glimpse(myreptiles_estimated$data)
# - `estimation_table` - a tibble with all unique size estimations that were joined to your data
dplyr::glimpse(myreptiles_estimated$size_estimations)
# - `model_table` - a tibble with all unique allometric models that were joined to your data
dplyr::glimpse(myreptiles_estimated$model_estimations)


# But now let's say you do not really want to do estimations on the data, but just see the results of all possible estimations. Fear not, it is easy!
## First compile the measurement table
measurement_table <-
  make_measurement_table(dats = myreptiles, ## The data to be used
                         level_vec  = level_vec) ## The taxonomic levels to be used
## Get all possible size estimations
size_estimation_table <-
  full_estimation_table(level_vec  = level_vec ,
                        measurement_table = measurement_table,
                        what = "size_col")
### Have a glimpse at it
dplyr::glimpse(size_estimation_table)

## Get all possible models
### Note that the package filters out models with p >0.05 and R2 < 0.95 to not include overfitted models
model_estimation_table <-
  full_estimation_table(level_vec  = level_vec ,
                        measurement_table = dry_wet(measurement_table,
                                                        biomass_type = "dry"), ## Filter for dry biomass
                        what = "biomass_col")
### Have a glimpse at it
dplyr::glimpse(model_estimation_table)

```

## Experimental feature: traits
Let's say that you do not have valid models under a taxonomic level that feels satisfying (e.g. order). Here one alternative would be to use traits instead of taxonomy to infer similarities between species. For example, two small, white, undefended worm-like creatures have a high likelihood of having similar body masses if of the same size.<br>
The package contains some functions (currently in development) to help you do so. So far, they are not incorporated into the `hellometry()` main function, but can get you as far as `full_estimation_table()`. To use them, you would need to supply in `measurement_table` a set of columns that contain <b>[fuzzy]( https://doi.org/10.1111/j.1365-2427.1994.tb01742.x)</b> traits. The functions will only work with fuzzy traits with discrete integer values, and will group species with those that have trait values +/-1 those of the focus species. You also tell the functions which column uniquely identifies a species through the `id_col` argument. Here is one example, adding a few random fuzzy traits to the reptile measurement table.
```{r}
# Fuzzy (discrete integer) trait columns
trait_columns <-
  c("t1", "t2", "t3")

# Add a few random fuzzy traits to the measurement table, one value per species
set.seed(1)
trait_lookup <-
  data.frame(species = unique(measurement_table$species))
for (tc in trait_columns)
  trait_lookup[[tc]] <- sample(1:3, nrow(trait_lookup), replace = TRUE)
dats <-
  dplyr::left_join(measurement_table, trait_lookup, by = "species")

## Get long format data of which species have similar traits
matched_traits <-
  matcher_of_traits(measurement_table = dats,
                    trait_columns = trait_columns,
                    id_col = "species")
### See the output
head(matched_traits)

## Get clean list for full_estimation_table, includes matcher_of_traits()
clean_trait_list <-
  make_trait_table(measurement_table = dats,
                   trait_columns = trait_columns,
                   id_col = "species")
### See the output
clean_trait_list[[1]]

# The last two functions are called in full_estimation_table if you add the trait arguments
## Size estimation table
size_estimation_table_traits <-
  full_estimation_table(level_vec  = c(),
                        measurement_table = dats,
                        traits = TRUE, ## traits = TRUE
                        trait_columns = trait_columns, ## vector of fuzzy trait columns
                        id_col = "species", ## column identifying species
                        what = "size_col")
### See the output
head(size_estimation_table_traits)

## Model table
model_estimation_table_traits <-
  full_estimation_table(level_vec  = c(),
                        measurement_table = dats,
                        traits = TRUE, ## traits = TRUE
                        trait_columns = trait_columns, ## vector of fuzzy trait columns
                        id_col = "species", ## column identifying species
                        what = "biomass_col")
### See the output
head(model_estimation_table_traits)
```
